Extracting The Power of Dependent Types

### 2021-07-09 paper.tex
Pass over abstract + introduction

@@

  Most existing programming languages provide little support to formally
  state and prove properties about programs.  Adding such capabilities is far
  from trivial, as it requires significant re-engineering of the existing
  compilers and tools.  This paper proposes a novel technique to write
  correct-by-construction programs in languages without built-in verification
  capabilities, maintaining the ability to use existing tools.
  This is achieved in three steps.  Firstly, we
  give a shallow embedding of the language (or a subset) into a
  dependently typed language.  Secondly, we write a program
  in that embedding, and we use dependent types to guarantee
  correctness properties of interest.  Thirdly, we
  translate the embedded program back into the original language,
  so that it can be used by the existing compilers and tools.

  Our main insight is that it is possible to
  express all three steps in a single language, as long as it supports
  both dependent types and reflection. Essentially,
  this allows us to express a program, its formal properties, and
  a compiler for it hand-in-hand, offering a lot of flexibility to programmers.
  We demonstrate this three-step approach by embedding a subset of
  the PostScript language in Agda, and illustrating it with several short examples.
  Thus we use the power of reflection to bring the benefits of dependent types to languages
  that had to go without them so far.

@@

\section{Introduction}

It is often desirable to guarantee that a certain class of errors does
not occur in the given program.  For example, the program can never dereference
a null pointer; or division by zero never occurs; or indexing is always
within bounds.  In practical programming languages, the kind of guarantees
that we can provide formally is typically very limited.  At the same time,
there could be many arguments in favour of keeping using these languages.
For example they may come with great tooling, compilers and libraries.
Therefore, the problem we are investigating in this paper is how to
provide formal guarantees for the program in some language, yet being
able to use the original compilers and tools.

One way to tackle this problem is via rephrasing it in terms
of dependent types.  If the language of interest were to support dependent
types, we could encode any property about the given program in
types.  Then, if the program typechecks, the property holds automatically.
So, the question now becomes how can we add dependent types to a language
(for verification part), and then eliminate them (for using the original
compiler).

While implementing a custom typechecker from scratch for a
dependently-typed language is possible, it is time-consuming and error-prone.
Alternatively, one can reuse existing implementations of dependently-type
systems by means of embedding.  In this case, we can share the type system
of the host language to typecheck the embedded language.

There are two common approaches to language embedding: deep and shallow.
Shallow embedding defines a language in terms of functions and operators
of the host language.  Programs in a shallowly-embedded language are programs
in the host language, so they are immediately runnable; types of the embedded
language are types of the host language, so the host typechecker is used for
checking embedded programs.  The downside of this approach is that we cannot
easily get the representation of the embedded program.

Deep embedding defines a data type to represent the AST of the embedded language.
Then we need to define a type system and evaluator for the language separately.
With a dependently-typed host language the AST and the type system can be
defined within a single data structure, which is typically referred as
tagless interpreters~\cite{}.  The benefit of deep embedding is that we
get immediate access to the structure of the embedded program.
The approach works beautifully for
embedded languages with simple type systems.  If the language we want to
embed is dependently-typed, it is shown~\cite{} to be possible, but the
resulting embedding is very difficult to use in practice.

Our solution to this problem is to rely on a reflection mechanism that
can be found in dependently-typed systems such as Idris, Agda, Lean.
Reflection makes it possible to obtain a representation of the
given term of the host language.  Essentially, shallow embedding in
a host language with reflections suddenly becomes deep.

In this paper we demonstrate that the proposed approach is actually possible
by example.  We embed a subset of PostScript~\cite{}, which is a stack-based
language used in the corresponding document format.  Simplicity of the language
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We use Agda as our host language, and we demonstrate the
embedding, how to specify a program, and most importantly, program extraction.

\todo[inline]{Should we talk about using rewrites as an optimisation?
        Should we talk here about scaling to realistic examples?
        Should we mention our work on extending Agda?}
\todo[inline]{Technical contributions}

@@

\section{Scaling to Bigger Langauges}
Not sure whether this deserves a separate section, but we need to explain
that the proposed technology perfectly scales to the level of bigger (more complex)
languages.

@@

\section{Conclusions and Future Work}

>>>>

\newcommand{\myref}[2]{\hyperref[#2]{#1~\ref*{#2}}}
\newcommand{\figref}[1]{\myref{Fig.}{fig:#1}}%{Fig.~\ref{fig:#1}}
\newcommand{\secref}[1]{\myref{Sect.}{sec:#1}}
\newcommand{\lemref}[1]{\myref{Lemma}{lem:#1}}
\newcommand{\thmref}[1]{\myref{Theorem}{thm:#1}}
\newcommand{\defref}[1]{\myref{Definition}{def:#1}}
\newcommand{\eqnref}[1]{\hyperref[#1]{(\ref{#1})}}

@@

  Most existing programming languages provide little support to formally
  state and prove properties about programs.  Adding such capabilities is far
  from trivial, as it requires significant re-engineering of the existing
  compilers and tools.  This paper proposes a novel technique to write
  correct-by-construction programs in languages without built-in verification
  capabilities, while maintaining the ability to use existing tools.
  This is achieved in three steps.  Firstly, we
  give a shallow embedding of the language (or a subset) into a
  dependently typed language.  Secondly, we write a program
  in that embedding, and we use dependent types to guarantee
  correctness properties of interest.  Thirdly, we
  extract a program written in the original language,
  so that it can be used by the existing compilers and tools.

  Our main insight is that it is possible to
  express all three steps in a single language that supports
  both dependent types and reflection. Essentially,
  this allows us to express a program, its formal properties, and
  a compiler for it hand-in-hand, offering a lot of flexibility to programmers.
  We demonstrate this three-step approach by embedding a subset of
  the PostScript language in Agda, and illustrating it with several short examples.
  Thus we use the power of reflection to bring the benefits of dependent types to languages
  that had to go without them so far.

@@

\section{Introduction}\label{sec:introduction}

When writing a program, it is often desirable to guarantee that a
certain class of errors does not occur in the given program.  For
example, the program can never dereference a null pointer; or division
by zero never occurs; or indexing is always within bounds.  In most
programming languages, the class of guarantees that can be checked
statically is very limited.  Nevertheless, moving to a different
language is not always an option.  For example, these languages may
come with great tooling, compilers and libraries.  In this paper, we
investigate how to provide strong static guarantees for programs in a
given language, while maintaining the ability to to use existing
compilers and tools.

One very powerful method for enforcing almost any functional property
of a program statically is by using \emph{dependent types}.
In a dependently typed language, properties can be encoded as types,
and if the program typechecks, the property is guaranteed to hold.
If we can extend our existing language with dependent types
(for verification part) and then eliminate them (for using the original
compiler), then our goal is achieved.

While it is possible to add dependent types to an existing language
from scratch, doing so is time-consuming and error-prone.  Instead,
one can also embed the language we want to use into an existing
dependently typed language, which allows us to re-use the typechecker
of the host language to typecheck the embedded language.

There are two common approaches to language embedding: deep and shallow.
A shallow embedding defines a language in terms of functions and operators
of the host language.  Programs in a shallowly-embedded language are programs
in the host language, so they are immediately runnable; types of the embedded
language are types of the host language, so the host typechecker is used for
checking embedded programs.  The downside of this approach is that we cannot
easily get the representation of the embedded program.

Meanwhile, a deep embedding defines a data type to represent the AST
of the embedded language.  This requires us to define a type system
and evaluator for the language separately.  With a dependently typed
host language the AST and the type system can be defined as a single
data structure, which is typically referred to as a tagless
interpreter~\cite{}.  The benefit of deep embedding is that we get
full access to the syntactic structure of the embedded program.  This
approach works beautifully for embedded languages with simple type
systems. However, for embedded languages with dependent types,
defining a tagless interpreter is technically possible~\cite{}, but
extremely challenging to define and use in practice.

We propose a novel approach that combines the best parts of shallow
and deep embeddings through the use of reflection in a
dependently-typed host language such as Coq, Agda, Idris, or Lean.
Concretely, we start by defining a shallow embedding of the target
language while making use of dependent types to statically enforce the
properties of interest, and write a program in this embedding. We then
make use of reflection to obtain a syntactic representation of the
program, and extract a program in the syntax of the target language
from it.  Finally, we compile the extracted program using existing
tools and libraries. Essentially, a shallow embedding in a host
language with reflections suddenly becomes deep.

In this paper we demonstrate that the proposed approach is actually possible
by example.  We embed a subset of PostScript~\cite{}, which is a stack-based
language used in the corresponding document format.  The simplicity of the language
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We use Agda as our host language, and we demonstrate the
embedding, how to specify a program, and most importantly, program extraction.

\paragraph{Contributions.}
\begin{itemize}

\item We propose a novel approach that can be used to statically
  enforce almost any property of programs in an existing language, by
  embedding it in a host language that supports dependent types and
  reflection.

\item We demonstate this approach by a concrete example: an embedding
  of (a subset of the) PostScript language into Agda.

\item We give several examples of PostScript programs written in our
  embedding, and demonstrate how to statically enforce properties such
  as the absence of stack underflows, termination, and functional
  correctness.

\item We show how to use reflection in Agda to implement an extractor
  that translates code written in our embedding to plain PostScript.

\item We show how to our approach provides easy support for partial
  evaluation by making use of normalisation in the host language. This
  allows us to use functions of the host language as macros, as well
  as automatically apply optimizations through the use of rewrite
  rules.
\end{itemize}

The rest of this paper is structured as follows. \secref{background}
provides background information on Agda and its reflection
capabilities. \secref{embedding} describes our embedding of PostScript
in Agda, and gives several examples of programs written in
it. \secref{extraction} shows how to implement an extractor for this
embedding using reflection. Next, \secref{partial-evaluation}
demonstrates how to apply partial evaluation to programs prior to
extraction by using Agda's normalisation procedure. Finally,
\secref{related} discusses related work and \secref{conclusion}
concludes.

@@

\section{Scaling to Bigger Languages} \label{sec:scaling}
Not sure whether this deserves a separate section, but we need to explain
that the proposed technology perfectly scales to the level of bigger (more complex)
languages.

@@

\section{Conclusions and Future Work} \label{sec:conclusion}
